---
title: "Data Wrangling"
author: "Rohan Kalu"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

# Data Wrangling

• Load necessary package
```{r}
library(dplyr)
library(magrittr)
library(ggplot2)
library(readr)
library(lubridate)
```

• Load the dataset
```{r}
accidents <- read.csv("data/accidents.csv")
guidance <- read.csv("data/guidance.csv")
```

• Check the first 6 data of the csv
```{r}
head(accidents)
```

• Check dimension, structure and summary of the data
```{r}
dim(accidents)

str(accidents)

summary(accidents)
```

### Data cleaning

• Change the data type of accident.date into 
```{r}
accidents <- accidents %>%
  mutate(Accident.Date = as.Date(Accident.Date, format="%d/%m/%Y"))

```

• Change Time..24hr. into 24 hr format
```{r}
accidents$`Time..24hr.` <- format(as.POSIXct(sprintf("%04d", accidents$`Time..24hr.`),
                                           format="%H%M", tz = "UTC"), format="%H:%M", usetz = FALSE)
head(accidents)
```
### EDA
• Check for missing values
```{r}
# Replace empty strings with NA in all columns
accidents[accidents == ""] <- NA
sapply(accidents, function(x) sum(is.na(x)))
```
• Look for the missing values
```{r}
missing_age <- accidents %>%
  filter(is.na(Age.of.Casualty))

# Filter rows with missing Daylight.Dark
missing_daylight.dark <- accidents %>%
  filter(is.na(Daylight.Dark))

missing_age
missing_daylight.dark
```

```{r}
# Convert `Daylight.Dark` to factor to handle missing values
accidents$Daylight.Dark <- factor(accidents$Daylight.Dark)

# Plotting using ggplot
ggplot(accidents, aes(x = factor(Lighting.Conditions), fill = is.na(Daylight.Dark))) +
  geom_bar(position = "stack") +
  labs(title = "Distribution of Missing Values in Daylight/Dark by Lighting Conditions",
       x = "Lighting Conditions",
       y = "Count",
       fill = "Missing Values") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue"),
                    labels = c("Missing", "Not Missing"),
                    guide = guide_legend(reverse = TRUE)) +
  theme_minimal()

```
### Data Exploration

• Check for anamolies
```{r}
unique(accidents$Road.Surface)
unique(accidents$X1st.Road.Class)
```

```{r}
accidents <- accidents %>%
  mutate(
    X1st.Road.Class = case_when(
      X1st.Road.Class %in% c("U") ~ "Unclassified",
      grepl("^A\\d+", X1st.Road.Class) ~ "A",
      grepl("^A\\(M\\)$", X1st.Road.Class) ~ "A(M)",
      grepl("^A\\d+\\(M\\)$", X1st.Road.Class) ~ "A(M)",
      grepl("^B\\d+", X1st.Road.Class) ~ "B",
      X1st.Road.Class %in% c("Motorway", "A", "B", "C", "Unclassified", "A(M)") ~ X1st.Road.Class,
      X1st.Road.Class %in% c("M62") ~ "Motorway",
      TRUE ~ as.character(X1st.Road.Class)
    ),
    `Road.Surface` = case_when(
      Road.Surface %in% c("Wet/Damp", "Wet \xa8 Damp", "Wet") ~ "Wet/Damp",
      Road.Surface == "Frost/Ice" ~ "Frost / Ice",
      Road.Surface == "Ice" ~ "Ice",
      Road.Surface == "Snow" ~ "Snow",
      Road.Surface == "1" ~ "Dry",
      Road.Surface == "2" ~ "Wet / Damp",
      Road.Surface == "3" ~ "Snow",
      Road.Surface == "4" ~ "Frost / Ice",
      Road.Surface == "5" ~ "Flood (surface water over 3cm deep)",
      TRUE ~ as.character(Road.Surface)
    ),
    `Lighting.Conditions` = case_when(
      Lighting.Conditions == 1 ~ "Daylight: street lights present",
      Lighting.Conditions == 2 ~ "Daylight: no street lighting",
      Lighting.Conditions == 3 ~ "Daylight: street lighting unknown",
      Lighting.Conditions == 4 ~ "Darkness: street lights present and lit",
      Lighting.Conditions == 5 ~ "Darkness: street lights present but unlit",
      Lighting.Conditions == 6 ~ "Darkness: no street lighting",
      Lighting.Conditions == 7 ~ "Darkness: street lighting unknown",
      TRUE ~ as.character(Lighting.Conditions)
    ),
    `Weather.Conditions` = case_when(
      Weather.Conditions == 1 ~ "Fine without high winds",
      Weather.Conditions == 2 ~ "Raining without high winds",
      Weather.Conditions == 3 ~ "Snowing without high winds",
      Weather.Conditions == 4 ~ "Fine with high winds",
      Weather.Conditions == 5 ~ "Raining with high winds",
      Weather.Conditions == 6 ~ "Snowing with high winds",
      Weather.Conditions == 7 ~ "Fog or mist – if hazard",
      Weather.Conditions == 8 ~ "Other",
      Weather.Conditions == 9 ~ "Unknown",
      TRUE ~ as.character(Weather.Conditions)
    ),
    `Casualty.Class` = case_when(
      Casualty.Class == 1 ~ "Driver or rider",
      Casualty.Class == 2 ~ "Vehicle or pillion passenger",
      Casualty.Class == 3 ~ "Pedestrian",
      TRUE ~ as.character(Casualty.Class)
    ),
    `Casualty.Severity` = case_when(
      Casualty.Severity == 1 ~ "Fatal",
      Casualty.Severity == 2 ~ "Serious",
      Casualty.Severity == 3 ~ "Slight",
      TRUE ~ as.character(Casualty.Severity)
    ),
    `Sex.of.Casualty` = case_when(
      Sex.of.Casualty == 1 ~ "Male",
      Sex.of.Casualty == 2 ~ "Female",
      TRUE ~ as.character(Sex.of.Casualty)
    )
  )
```

• Check for anamolies again
```{r}
unique(accidents$X1st.Road.Class)
unique(accidents$Road.Surface)
```

```{r}
head(accidents)
```
```{r}
unique(accidents$Weather.Conditions)
```
• Check for anamolies on other columns
```{r}
exclude_cols <- c("Age.of.Casualty", "Accident.Date", "Time..24hr.")

# Select columns except Age, Date, and Time
selected_cols <- accidents[, !names(accidents) %in% exclude_cols]

# Check unique values for each selected column using sapply
unique_values <- sapply(selected_cols, function(x) length(unique(x)))

# Print the unique values for each column
unique_values
```

```{r}
accidents <- accidents %>% 
  select(-Local.Authority, -Daylight.Dark)
```
```{r}
colnames(accidents)
```

### Outliers Detection

• Use Boxplot to detect outliers
```{r}
filtered_data <- accidents[!is.na(accidents$Age.of.Casualty), ]

# Visual inspection with boxplot
ggplot(filtered_data, aes(y = Age.of.Casualty)) +
  geom_boxplot()
```

• Use 3-Sigma method
```{r}
age_of_casualty <- accidents$Age.of.Casualty
mean_age = mean(age_of_casualty, na.rm = TRUE)
sd_age = sd(age_of_casualty, na.rm = TRUE)
upper_bound <- mean_age + 3 * sd_age
lower_bound <- mean_age - 3 * sd_age
outliers_dataset <- accidents %>%
  filter(Age.of.Casualty < lower_bound | Age.of.Casualty > upper_bound)
number_of_outliers_sigma <- nrow(outliers_dataset)
print(paste("Number of outliers:", number_of_outliers_sigma))
print(outliers_dataset)
```

• IQR
```{r}
# Calculate interquartile range (IQR) for Age_of_Casualty
Q1 <- quantile(filtered_data$Age.of.Casualty, 0.25, na.rm = TRUE)
Q3 <- quantile(filtered_data$Age.of.Casualty, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

# Calculate lower and upper bounds for outliers
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify outliers
outliers <- filtered_data[filtered_data$Age.of.Casualty < lower_bound | filtered_data$Age.of.Casualty > upper_bound, ]

# Print count of outliers
outliers  # or use dim(outliers)[1] for the number of rows
```

• Hampel Identifier
```{r}
# Extract Age of Casualty data
age_casualty <- filtered_data$Age.of.Casualty

# Calculate the median and MAD
median_age <- median(age_casualty, na.rm = TRUE)
mad_age <- mad(age_casualty, na.rm = TRUE)

# Calculate Hampel's upper and lower bounds
hampel_upper_bound <- median_age + 3 * mad_age
hampel_lower_bound <- median_age - 3 * mad_age

# Identify outliers using Hampel identifier
outliers_hampel <- age_casualty[age_casualty > hampel_upper_bound | age_casualty < hampel_lower_bound]

# Print the outliers
print(outliers_hampel)

# Create a dataset for Hampel outliers
outliers_hampel_dataset <- filtered_data %>%
  filter(Age.of.Casualty > hampel_upper_bound | Age.of.Casualty < hampel_lower_bound)

# Print the number of Hampel outliers
number_of_outliers_hampel <- nrow(outliers_hampel_dataset)
print(paste("Number of Hampel outliers:", number_of_outliers_hampel))

# Print the Hampel outliers dataset
print(outliers_hampel_dataset)

```

```{r}
write.csv(filtered_data, file = "clean_accident.csv", row.names = FALSE)
```


# Data Exploration

• Load necessary packages
```{r}
# Note: Please run data Wrangling first
# Unknown error preventing this rmd file to be converted into html or pdf
library(ggplot2)
library(magrittr)
library(readr)
```

• Load the cleaned csv
```{r}
data <- read_csv("clean_accident.csv")
```
• Check the first 6 rows 
```{r}
head(data)
```

• Check the structure of the data
```{r}
str(data)
```

```{r}
male_accidents <- accidents %>%
  filter(Casualty.Class == "Driver or rider", Sex.of.Casualty == "Male")

female_accidents <- accidents %>%
  filter(Casualty.Class == "Driver or rider", Sex.of.Casualty == "Female")

count(male_accidents)
count(female_accidents)
```

• Group the data by weather
```{r}
male_counts <- male_accidents %>%
  group_by(Weather.Conditions) %>%
  summarise(total_male_accidents = n())

female_counts <- female_accidents %>%
  group_by(Weather.Conditions) %>%
  summarise(total_female_accidents = n())
```

```{r}
accident_comparison <- merge(male_counts, female_counts, by = "Weather.Conditions", all = TRUE)

# Calculate the difference where males have more accidents than females
accident_comparison <- accident_comparison %>%
  mutate(more_male_accidents = ifelse(total_male_accidents > total_female_accidents,
                                      total_male_accidents - total_female_accidents,
                                      0))

# Filter for cases where males have more accidents than females
more_accidents <- accident_comparison %>%
  filter(more_male_accidents > 0) %>%
  arrange(desc(more_male_accidents))  # Sort by descending difference
```

```{r}
# Create a bar plot using ggplot2
ggplot(more_accidents, aes(x = reorder(Weather.Conditions, more_male_accidents), y = more_male_accidents)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  labs(x = "Weather Conditions", y = "More Male Accidents") +
  ggtitle("Weather Conditions Where Males Have More Accidents Than Females") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
number_of_cases <- nrow(more_accidents)
number_of_cases
```

```{r}
accidents <- accidents %>%
  mutate(Year = lubridate::year(Accident.Date))

# Group by year and count casualties
casualties_by_year <- accidents %>%
  group_by(Year) %>%
  summarise(total_casualties = n())

# Find the year with the highest number of casualties
max_casualty_year <- casualties_by_year %>%
  filter(total_casualties == max(total_casualties)) %>%
  pull(Year)

# Plotting the trend of casualties over the years
ggplot(casualties_by_year, aes(x = Year, y = total_casualties)) +
  geom_line() +
  geom_point() +
  labs(title = "Trend of Casualties Over Years",
       x = "Year",
       y = "Total Casualties") +
  theme_minimal()

# Print the year with the highest number of casualties
print(paste("Year with the highest number of casualties:", max_casualty_year))
```

### Light condition and Severity
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Assuming 'accidents' dataset is already loaded and cleaned

# Create the bar plot for Light Conditions and Severity
light_severity_plot <- ggplot(accidents, aes(x = Lighting.Conditions, fill = Casualty.Severity)) +
  geom_bar(position = "dodge") +  # Create a bar plot with bars side by side for each severity level
  labs(title = "Casualty Severity by Lighting Conditions",
       x = "Lighting Conditions",
       y = "Count of Casualties",
       fill = "Casualty Severity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability

# Print the plot
print(light_severity_plot)

```

```{r}
# Check the count of casualties by lighting conditions and severity
table(accidents$Lighting.Conditions, accidents$Casualty.Severity)

```

##### Interpretation
###### Darkness with no streetlight:
There is no fatal injury and only few seriously injured with a slightly higher number minorly injured

###### Darkness: street lighting unknown
Although no fatal injuries, the number of seriously and slightly injured has tripled compared to darkness with no streetlight.

###### Darkness: street lights present and lit
The lit streetlight at darkness has the highest number of fatal injuries (15), 107 seriously injured and 405 slightly injured. The reason might be because of low volume of traffic at night and well lit streets.

###### Darkness: street lights present but unlit
The unlit streetlight in darkness has the lowest number of injuries overall(0 fatal, 1 serious, 3 slight). The reason might be because of although low traffic, but low visibility as well.

###### Daylight: street lights present
Shows the highest number of accidents in every category except fatality(10 fatal, 178 serious, 1263). The plausible reason might be high volume of traffic and better visibility of accidents happening on the broad daylight.

###### Other 2 conditions (Daylight: no street lights present and Daylight: streetlight unknown) has 0 number of casualties as the obvious reason being no streetlight being lit during the daytime.

### Weather and number of vehicles involved
```{r}
weather_vehicles_count_data <- accidents %>%
  group_by(Weather.Conditions) %>%
  summarise(Count_of_Accidents = n())

weather_vehicles_plot <- ggplot(weather_vehicles_count_data, aes(x = Weather.Conditions, y = Count_of_Accidents)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Accidents by Weather Conditions",
       x = "Weather Conditions",
       y = "Count of Accidents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

weather_vehicles_plot
```
##### Interpretation

###### Fine without high winds:
The weather condition has the highest number of casualties reaching almost 2000 and is exponentially higher compared to the next in line (Raining without high winds). The likely reason might be high volume of traffic

###### Raining without high winds:
Although not as severe as fine without high winds, the number of vehicles involved in accident is noticeable (250). The most probable reason cold be more cautious driving during rain as the tires may slip.

###### Raining with high winds:
The third condition with the highest number of vehicles involved in accident but incomparable to aforementioned ones (100). The probable reason could be even more cautios driving than the rain without winds as there are 2 factors to look at.

###### Others (everything except aforementioned):
All the other weather conditions have less than 50 vehicles involved in an accident.
```{r}
table(accidents$Weather.Conditions, accidents$Number.of.Vehicles)
```

# Linear Regression

• Load necessary libraries
```{r}
library(dplyr)
library(magrittr)
library(tidyverse)
```

• Import clean accident csv
```{r}
data <- read.csv('data/accidents.csv')
head(data)
```

• Check the structure of data
```{r}
str(data)
```

• Convert columns into something usable
```{r}
data$Casualty.Class <- as.factor(data$Casualty.Class)
data$Casualty.Severity <- as.factor(data$Casualty.Severity)
data$Type.of.Vehicle <- as.factor(data$Type.of.Vehicle)
data$Weather.Conditions <- as.factor(data$Weather.Conditions)
str(data)
```
• Create a data to train
```{r}
train.data <- data
head(train.data)
```


• Create a linear model
```{r}
lm.model <- lm(Age.of.Casualty ~ Casualty.Class + Casualty.Severity 
               + Type.of.Vehicle + Weather.Conditions, data = train.data)

summary(lm.model)
```
• Filter data with missing values
```{r}
missing.data <- data %>%
  filter(is.na(Age.of.Casualty))
```

• Predict the missing values
```{r}
predicted.age <- predict(lm.model, newdata = missing.data)

data$Age.of.Casualty[is.na(data$Age.of.Casualty)] <- predicted.age

missing.data <- data %>%
  filter(is.na(Age.of.Casualty))
missing.data
```

• Round up the predicted value and convert it to integer
```{r}
data$Age.of.Casualty <- round(data$Age.of.Casualty)

data$Age.of.Casualty <- as.integer(data$Age.of.Casualty)
```

• Check for any remaining missing values in the dataset
```{r}
missing.values <- colSums(is.na(data))
print(missing.values[missing.values > 0])
```
```{r}
dim(data)
```

```{r}
write.csv(data, "regression.csv", row.names = FALSE, quote= FALSE, fileEncoding = "UTF-8")
```


